---
title: "PSET 10"
author: "Sarah Van Alsten"
date: "4/6/2020"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plm)
library(lfe)

load("wagepan_wide.RData")
```

## Question 1

a.
```{r}
#make a vector of column names
names <- names(wage_wide)[-1]

#reshape data to long
dat <- wage_wide %>% 
  pivot_longer(cols = all_of(names),
               names_to = c("characteristic", "year"),
               values_to = "value",
               names_sep = "_") %>%
  pivot_wider(id_cols = c("nr", "year"), names_from = "characteristic",
              values_from = "value")

```

b.
$\delta_t$ is a time fixed effect with an impact on all units (common shock)

$\eta_i$ is a unit-level fixed effect

$\epsilon_{it}$ is the time varying error (error for unit $i$ at time $t$)


c.
```{r}
#set seed to randomly pick workers
set.seed(02139)

#sample 10 workers 
ids <- sample(unique(dat$nr), size = 10, replace = FALSE)

#select these workers and plot results
dat %>%
  filter(nr %in% ids) %>%
  ggplot(aes(x = year, y = lwage, color = factor(nr), group = factor(nr))) +
  geom_path() + theme_bw() +
  labs(x = "Year", y = "Log Hourly Wage", color = "ID")


```

In general, there isn't a clear pattern other than that units ended with higher wages than they began with and that trends are not linear/constant over time. The variation within units seems to be slightly larger than the variation between units, as the spread of wages seems less pronounced than the variability over time within an observation.

d.

$\mathbb{V}(Y_i) = \mathbb{E}[Var(Y)|X] + Var(\mathbb{E}[Y|X])$

$Var(\mathbb{E}[Y|X])$ is the between-unit variance, and refers to how much difference there is from one unit to another, or how much the differences between each individual's wages explains the total variation in wages over time.

$[Var(Y)|X]$ is the within-unit variance, and refers to how much variability any one individual sees in their own wages over time.

```{r}
#E{Var(Y)|X} = mean of var within
var.within <- dat %>%
  group_by(nr) %>%
  summarise(group.var = var(lwage, na.rm = T)) %>%
  summarise(e.group.var = mean(group.var))

#var total
var.total <- var(dat$lwage, na.rm = T)


#var btwn = total - within
var.btwn <- var.total - var.within

#print results
print(paste("Variance Within:", sprintf("%.4f", var.within[1,1])))
print(paste("(Mean) Variance Between:", sprintf("%.4f", var.btwn)))
print(paste("Variance Total:", sprintf("%.4f", var.total)))
```

The between-unit variance is slightly larger than the within-unit variance, which is consistent with the graph.
e.
```{r}
#make this manually so the name of the var is consistent in stargazer part f/g
dat$exper_sq <- dat$exper^2

#fit model assuming uncorrelated error
mod <- lm(lwage ~ educ + black + hisp + exper +
           exper_sq+ married + union, data = dat)
summary(mod)

#make nice output
stargazer::stargazer(mod, float = FALSE)

```

The effect of belonging to a union is an increase of 0.180 (se = 0.017) on log of wages.
We are estimating the pooled effect - the effect of union membership on wages assuming that the effects of all other covariates do not change over time and that there is no serial autocorrelation for units. In other words, that all the variation in wages not capture by our model does not depend on the error at different periods in time (and hence why we only include $\epsilon_{it}$ and not $\alpha_i$. )
Because we assumed no autocorrelation in order to say that the estimate was unbiased, we don't need to cluster the standard errors- the potential outcomes for each unit $i$ at each time $t$ are assumed to not depend on one another. Thus, conventional OLS SEs are okay given our assumptions.

f.
```{r, warning=FALSE, message=FALSE}
#make de-meaned variables

#copy data set so can use mutate all without messing up raw data
dat.dm <- dat

#make demeaned versions of all predictors + y variable
dat.dm <- dat.dm %>%
  group_by(nr) %>%
  select(-year)%>%
  mutate_all(.funs = ~(. - mean(., na.rm = T))) 


#add year back on            
dat.dm$year <- dat$year

#estimated demeaned model (all regressors are their demeaned selves)
mod.dm <- lm(lwage ~ educ + black + hisp + exper +
            exper_sq + married + union -1, data = dat.dm) 

#correct total degrees of freedom= Nt- N - K
mod.dm$df.residual <- (8*545 - 545 - 4)
summary(mod.dm) 
         

#report results
stargazer::stargazer(mod.dm, float = FALSE)

```

Time invariant variables (education, Black, and Hispanic) are dropped because they do not contribute to within-unit wage changes (the rank condition states that predictors must change overtime for at least some units, which is not the case for these characteristics). The estimated effect of union membership is much smaller in the within model than the pooled model (0.084 vs 0.180).

For this model, we are assuming that there is no serial correlation (the wages in one time period aren't dependent on the wages for an earlier period), that the each unit's treatment doesn't affect the potential outcomes of other units (no interference), that the effect of union membership is the same for all treated units (no different version of treatment), that higher wages don't increase probability of being in a uninon (simultaneity), that we have not omitted any time-varying predictors of the outcome (e.g. a gradual roll out of a minimum wage increase affecting only some workers), and, relatedly, that there is exogeneity conditional on the any unobserved effect. It seems highly unlikely that we don't observe serial correlation - an individuals' wages are generally partially dependent upon previous wages, such as when they get a salary increase, or would only accept a new job with higher pay than their current position. This implies that some of what our error term is capturing - like an employee's perceived merit by their employer; which would affect wages - is persistent over time and errors would be correlated.


g.
```{r, warning=FALSE, message = FALSE, error=FALSE}

#estimate with plm
pdata <- pdata.frame(dat, c("nr", "year"))
mod.fe <- plm(lwage ~ educ + black + hisp + exper +
                exper_sq+ married + union, data = pdata, index = nr) 
#summary(mod.fe)
#clustered Std Errors accounting for serial correlation
ses <- lmtest::coeftest(mod.fe, vcov=function(x) vcovHC(x, cluster="group", type="HC1"))
#show results in a table
stargazer::stargazer(mod.dm, mod.fe,
                     se = list(summary(mod.dm)$coef[,2], ses[,2]),
                     float = FALSE)


# #estimate with lfe, clustering SEs on unit
# mod.fe <- felm(lwage ~ educ + black + hisp + exper +
#                      exper_sq + married + union | nr | 0 |nr, data = dat)
# #summary(mod.fe)
# 
# #show results in a table
# stargazer::stargazer(mod.dm, mod.fe, 
#                      float = FALSE)

```

All point estimates are the same, although the standard errors differ because we accounted for clustering.

h.
```{r, warning=FALSE, message = FALSE}

mod.tf <- plm(lwage ~ educ + black + hisp + exper +
                exper_sq+ married + union, data = pdata, 
              index = nr, effect = "twoways", model = "within") 
summary(mod.tf)
#clustered Std Errors accounting for serial correlation
ses.tf <- lmtest::coeftest(mod.tf, vcov=function(x) vcovHC(x, cluster="group", type="HC1"))


# #estimate with lfe, clustering SEs on unit
# mod.tf <- felm(lwage ~ educ + black + hisp + exper +
#                      exper_sq + married + union | nr + year| 0 |nr, data = dat)
# summary(mod.tf)

#print out results
stargazer::stargazer(mod.tf, float = FALSE, se = ses.tf)

```

In this model compared to the previous, the estimated effect of unions decreases slightly (0.080 vs 0.082), while the coefficients for married and experience squared increase in magnitude. Time fixed effects accounts for common shocks that affect all units in each time period.
The variables education, Black, Hispanic, and experience all dropped out. Education, Black, and Hispanic drop out for the same reason as in the previous model: they do not change at all through time and thus do not contribute to within-unit variation. Experience drops out because it is a linear function of the time variable (each additional year = 1 point increase in experience) so it is perfectly collinear with time and doesn't add anything additional to the model.

i.
```{r}
#calculate # units with 0 variance in union status
dat %>%
  group_by(nr) %>%
  #sum of years belonging to a union
  summarise(union.var = sum(union)) %>%
  ungroup() %>%
  #total with 0 variance
  summarise(sum(union.var == 0 | union.var == 8))


```

Units with no/little variation in union status generally should increase the standard errors of the estimate for union, because they are adding to the degrees of freedom but not contributing much variation to the estimate. Unlike those that do have some variation, the estimated effect for units with no variation is calculated using only the variation in wages over time, since there is no variation within the individual.

j.
```{r}
#linear trend 
dat$year_num <- as.numeric(pdata$year)
pdata$year_num <- as.numeric(pdata$year)

# #estimate with lfe, clustering SEs on unit
# mod.lt <- felm(lwage ~ educ + black + hisp + exper +
#                      exper_sq + married + union + year_num | nr | 0 |nr, data = dat)
# summary(mod.lt)


#include unit specific linear time effect in model
mod.time.trend <- plm(lwage ~ educ + black + hisp + exper +
                      exper_sq + married + union + factor(nr)*year_num,
                      data = pdata,
                      effect = "twoways",
                      model = "within")

#coefficient for union membership
mod.time.trend$coefficients[3]
ses.tt <- lmtest::coeftest(mod.time.trend, vcov=function(x) vcovHC(x, cluster="group", type="HC1"))

#show the se for union specifically
ses.tt[,2][[3]]

```

k.

```{r}
#plot individual 408's wages by year
dat %>%
  filter(nr == 408) %>%
  ggplot(aes(x = year, y = lwage, color = factor(union),
             group = nr)) + geom_path() + geom_point() +
  theme_classic() + labs(x = "Year", y = "Log Wages", color = "In Union")
```

```{r}
#run 3 diff models
#unit fixed effects
u.fe <- felm(lwage ~ union | nr | 0 |nr,
             data = dat)

#unit and time fixed effects
ut.fe <- felm(lwage ~ union | nr + year| 0 |nr,
             data = dat)


#unit fixed effects and time linear trend
u.fe.lt <- felm(lwage ~ union + year_num| nr| 0 |nr,
               data = dat)

#lapply(list(u.fe, ut.fe, u.fe.lt), FUN = summary)

#add predicted values to the data frame
dat$u.fe.pred <- u.fe$fitted.values
dat$ut.fe.pred <- ut.fe$fitted.values
dat$u.fe.lt.pred <- u.fe.lt$fitted.values


#plot individual 408's wages by year
dat %>%
  filter(nr == 408) %>%
  ggplot(aes(x = year, y = lwage, group = nr)) + geom_point() +
  theme_classic() + labs(x = "Year", y = "Log Wages", color = "Model") +
  geom_path(aes(x = year, y = u.fe.lt.pred, color = "Unit Fixed Effects + Linear Time Trend")) +
    geom_path(aes(x = year, y = u.fe.pred, color = "Unit Fixed Effects")) +
    geom_path(aes(x = year, y = ut.fe.pred, color = "Unit + Time Fixed Effects")) 
  

```


l.

No, I would disagree with him.