---
title: "PSET 5"
author: "Sarah Van Alsten"
date: "2/18/2020"
output:
  pdf_document:
        latex_engine: lualatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = 'center')

#open libraries
library(tidyverse)
library(Matching)

```


## Question 1. The Curse of Dimensionality

a. General Expression to Calculate Euclidean Distance between Two Points:

$$Distance = \sqrt{(x_{i1} - x_{j1})^2 + (x_{i2} - x_{j2})^2 ... + (x_{iP} - x_{jP})^2}$$
```{r}
#write a function that will give us the distance for a given # of covariates
getEuclidean <- function(numCov){
  
  #take the numCov columns to compare to
  compareDat <- as.data.frame(dat[, 1:numCov])
  
  
  #compute euclidean distance for each obs
  compareDat <- compareDat %>%
    #squared distance from 0 for each column
    mutate_all(.funs = ~((0 - .)^2)) %>%
    #create new column as sqrt of sum of these squared distances
    mutate(euclidean = sqrt(rowSums(.)))
  
  #return the minimum euclidean distance
  return(min(compareDat$euclidean))
  
}

```



b. 

```{r, fig.cap = 'Euclidean Distance by Covariate Number'}
#Generate dataset X of 500 obs, 20 covariates (normally distributed 0-1)
set.seed(02139)
dat <- as.data.frame(replicate(n = 20, expr = rnorm(n = 500, mean = 0, sd = 1)))

#make a blank dataframe to hold results
euc.res <- as.data.frame(cbind(1:20, rep(NA, 20)))

#get euclidean distance for 1 covariate
getEuclidean(numCov = 1)

#add this to the result data
euc.res$V2[1] <- getEuclidean(numCov = 1)

#now do this for the 2:20 covariates
for (i in 2:20){
  euc.res$V2[i] <- getEuclidean(numCov = i)
}

#plot results
euc.res %>%
  ggplot(aes(x = V1, y = V2)) +
  geom_point() +
  geom_path() +
  theme_bw() +
  labs(x = "Number of Covariates",
       y = "Minimum Euclidean Distance") +
  ggtitle("Euclidean Distance by Number Of Covariates")


```

c. These results demonstrate that as we add more dimensions or covariates to match on, the dissimilarity (distance) between even the best match and the index obervation increases (i.e. it becomes more difficult to find a very close/similar match).

## Question 2

```{r, echo = FALSE}
#read in the data
nsw <- haven::read_dta("C://Users//Owner//OneDrive//Documents//Spring2020//Causal_Inf//psetdata//Causal_PSETs//data//nsw_exper.dta")


psid <- haven::read_dta("C://Users//Owner//OneDrive//Documents//Spring2020//Causal_Inf//psetdata//Causal_PSETs//data//nsw_psid.dta")


```


a. 

```{r}
#unbiased estiamte of ATE
nsw.t <- t.test(re78 ~ nsw, data = nsw)

#ate and se
(ate.t <- nsw.t$estimate[2] - nsw.t$estimate[1])
(stderr.t <- nsw.t$stderr)


#re-estimate using linear regression
mod.nsw <- lm(re78 ~ nsw + age + educ + black + hisp + married + re74 + u74,
              data = nsw)

#get results
summary(mod.nsw)

#account for randomziation, use robust SE
sqrt(diag(sandwich::vcovHC(mod.nsw, type="HC2")))


```

The unadjusted estimate of the average treatment effect of NSW participation on 1978 earnings was an increase of \$1794 dollars (se = \$671). After adjusting for age, race, ethnicity, education, marital status, earnings in 1974, and employment in 1974, the average treatment effect of NSW participation was somewhat lower (increase of \$1721, se = \$678). This suggests that there may have been some confounding by the covariates, but not a substantial amount.

b.

```{r}
#calculate naive ATE using non-experimental data
psid.t <- t.test(re78 ~ nsw, data = psid)

#ate and se
(ate.psid <- psid.t$estimate[2] - psid.t$estimate[1])
(stderr.psid <- psid.t$stderr)

#re-estimate using regression
mod.psid <- lm(re78 ~ nsw + age + educ + black + hisp + married + re74 + u74,
              data = psid)

#get results
summary(mod.psid)

#account for randomziation, use robust SE
sqrt(diag(sandwich::vcovHC(mod.psid, type="HC2")))


```

We are estimating the average treatment effect conditional on all the covariates (age, education, race, ethnicity, marital status, earnings in 1974, and unemployment in 1974). The only difference is that now, we are using data from the general population, so we don't expect these variables to be balanced.
These methods do not recover the experimental results because, even conditional on all the covariates we adjusted for, the treated (those who were given the work program) and untreated (the general population) are not exchangeable (we don't have conditional ignorability). This indicates lack of balance on other unobserved/not included covariates, or that treated and control units do not tend to have overlap in covariate values, so we lack positivity when conditioning on these.


c.
```{r}
#estimate the propensity scores

mb  <- MatchBalance(nsw ~ age + educ + black +
                    hisp + married + re74 + re75  +
                    u74 + u75 + u78,
                    data = psid, 
                    nboots = 10)



#make a nice balance table
btab <- tableone::CreateTableOne(vars = c("age", "educ", "black",
                    "hisp" , "married", "re74" , "re75", "u74", "u75", "u78"),
                    data = psid,
                    strata = "nsw",
                    factorVars = c("black",
                    "hisp", "married", 
                    "u74", "u75" , "u78"))

#also get the K-S tests for distributions from the match balance
#pvalues for the KS tests for all variables
ks.pval <- c(NA,
             mb$BeforeMatching[[1]][9][[1]][[1]][[1]],
             mb$BeforeMatching[[2]][9][[1]][[1]][[1]],
             "--", #black is categorical
             "--", #hisp is categorical
             "--", #married is categorical
             mb$BeforeMatching[[6]][9][[1]][[1]][[1]],
             mb$BeforeMatching[[7]][9][[1]][[1]][[1]],
             "--", #u74 is categorical
             "--", #u75 is categorical
             "--") #u78 is categorical

#replace 0 with "<0.001"
ks.pval <- str_replace(ks.pval, pattern = "0", replacement ="< 0.001")


#print balance table with KS pvalues
as.data.frame(cbind(print(btab), ks.pval)) %>%
  kableExtra::kable() %>%
  kableExtra::kable_styling("striped")

                    
```

Based on the balance table, all covariates except for Hispanic ethnicity differed between the treated and control groups. In particular, unemployment in 1974 and 1975, being Black, being unmarried, and having low real earnings in 1974 and 1975 were most strongly associated with treatment status.

d.
```{r, fig.cap = 'Propensity Score Distributions'}
#estimate propensity scores using logistic regression in experimetn
pscore_model_exp <- glm(nsw ~ age + educ + black +
                    hisp + married + re74 + re75  +
                    u74 + u75 + u78,
                    data = nsw, 
                    family = binomial(link = logit))


pscore_exp <- predict(pscore_model_exp, type = "response")

#plot distributions of pscores
nsw$pscore <- pscore_exp

nsw %>%
  ggplot(aes(x = pscore_exp, fill = factor(nsw),alpha = .9)) +
  geom_density() +
  scale_fill_discrete(name = "NSW")+
  scale_alpha_continuous(guide = F) +
  ggtitle("Propensity Scores Experimental") +
  theme_bw()


#estimate propensity scores using logistic regression in non-experiment
pscore_model <- glm(nsw ~ age + educ + black +
                    hisp + married + re74 + re75  +
                    u74 + u75 + u78,
                    data = psid, 
                    family = binomial(link = logit))


pscore <- predict(pscore_model, type = "response")

#plot distributions of pscores
psid$pscore <- pscore

psid %>%
  ggplot(aes(x = pscore, fill = factor(nsw), alpha = .3)) +
  geom_density() +
  scale_fill_discrete(name = "NSW")+
  scale_alpha_continuous(guide = F) +
  ggtitle("Propensity Scores Non-Experimental") +
  theme_bw()

#to see better, also plot separately
psid %>%
  ggplot(aes(x = pscore, fill = factor(nsw), alpha = .3)) +
  geom_density() +
  scale_fill_discrete(name = "NSW") +
  facet_wrap(~nsw, scales = "free")  +
  theme_bw() +
  scale_alpha_continuous(guide = F) 

```

In the experimental data, there is substantial overlap in the distributions of propensity scores - those who got treatment seem to have slightly more observations with probability of treatment greater than 0.5 than the control, but otherwise they are very similar.

Conversely, there is very little overlap between the two propensity score distributions in the non-experimental data. Propensity scores in the control units are heavily skewed towards 0, and those in the treatment units are skewed towards 1. This indicates that control units had almost no probability of receiving treatment, whereas a large portion of those in the treatment group had almost complete probability of receiving treatment.

The two differ because the process of randomization in the experimental data helped ensure balance between the two groups AND similar propensity scores (if propensity = probability of treatment and treatment was randomly assigned with equal probability, then propensity distributions should also be comporable), whereas in non-experimental data, pre-treatment covariates were not balanced, especially given the nature of the the treatment (a training program is most likley offered to those with low employment or job skills)


e.
```{r}

#make matches
match.mah <- Match(Y = psid$re78,
                   Tr = psid$nsw,
                   X = psid[, c("age", "black", "u74", "educ", "married")],
                   M = 1, estimand="ATT",
                   Weight = 2)

summary(match.mah)

#check balance on all covariates (matched + unmatched)
mb2 <- MatchBalance(nsw ~ age + educ + black + u74 +
             married + hisp + re74 + re75  +
             u75 + u78,
             data = psid, 
             match.out = match.mah,
             nboots = 10)

#vector of covariates after matching
after.tr <- c(mb2$AfterMatching[[1]][3],
           mb2$AfterMatching[[2]][3],
           mb2$AfterMatching[[3]][3],
           mb2$AfterMatching[[4]][3],
           mb2$AfterMatching[[5]][3],
           mb2$AfterMatching[[6]][3],
           mb2$AfterMatching[[7]][3],
           mb2$AfterMatching[[8]][3],
           mb2$AfterMatching[[9]][3],
           mb2$AfterMatching[[10]][3])

#vector of values in control after matching
after.co <- c(mb2$AfterMatching[[1]][4],
           mb2$AfterMatching[[2]][4],
           mb2$AfterMatching[[3]][4],
           mb2$AfterMatching[[4]][4],
           mb2$AfterMatching[[5]][4],
           mb2$AfterMatching[[6]][4],
           mb2$AfterMatching[[7]][4],
           mb2$AfterMatching[[8]][4],
           mb2$AfterMatching[[9]][4],
           mb2$AfterMatching[[10]][4])

#sd in treated and control
var.tr <- c(mb2$AfterMatching[[1]][5],
           mb2$AfterMatching[[2]][5],
           mb2$AfterMatching[[3]][5],
           mb2$AfterMatching[[4]][5],
           mb2$AfterMatching[[5]][5],
           mb2$AfterMatching[[6]][5],
           mb2$AfterMatching[[7]][5],
           mb2$AfterMatching[[8]][5],
           mb2$AfterMatching[[9]][5],
           mb2$AfterMatching[[10]][5])

var.co <- c(mb2$AfterMatching[[1]][6],
           mb2$AfterMatching[[2]][6],
           mb2$AfterMatching[[3]][6],
           mb2$AfterMatching[[4]][6],
           mb2$AfterMatching[[5]][6],
           mb2$AfterMatching[[6]][6],
           mb2$AfterMatching[[7]][6],
           mb2$AfterMatching[[8]][6],
           mb2$AfterMatching[[9]][6],
           mb2$AfterMatching[[10]][6])

#make numeric and get sqrts
var.tr <- as.numeric(var.tr)
var.co <- as.numeric(var.co)

sd.tr <- sqrt(var.tr)
sd.co <- sqrt(var.co)


#vector of pvalues comparing groups
match.p <- c(mb2$AfterMatching[[1]][7],
           mb2$AfterMatching[[2]][7],
           mb2$AfterMatching[[3]][7],
           mb2$AfterMatching[[4]][7],
           mb2$AfterMatching[[5]][7],
           mb2$AfterMatching[[6]][7],
           mb2$AfterMatching[[7]][7],
           mb2$AfterMatching[[8]][7],
           mb2$AfterMatching[[9]][7],
           mb2$AfterMatching[[10]][7])

#also p values for KS tests
ks.pval2 <- c(mb2$AfterMatching[[1]][9][[1]][[1]][[1]],
             mb2$AfterMatching[[2]][9][[1]][[1]][[1]],
             "--", #black is categorical
             "--", #u74 is categorical
             "--", #married is categorical
             "--", #hisp is categorical
             mb2$AfterMatching[[7]][9][[1]][[1]][[1]], #re74
             mb2$AfterMatching[[8]][9][[1]][[1]][[1]], #re75
             "--", #u75 is categorical
             "--") #u78 is categorical

#make format nicer
ks.pval2 <- sprintf(as.numeric(ks.pval2), fmt = "%.3f")

#bind cols to make table
df <- as.data.frame(cbind(sprintf(after.tr, fmt = "%.3f"),
                    sprintf(after.co, fmt = "%.3f"),
                    sprintf(sd.tr, fmt = "%.3f"),
                    sprintf(sd.co, fmt = "%.3f"),
                    sprintf(match.p, fmt = "%.3f"),
                    ks.pval2)) 

#add labels
names(df) <- c("mean.tr", "mean.co",
               "sd.tr", "sd.co",
               "pvalue", "KSpvalue")

rownames(df) <- c("age", "educ", "black", "u74",
                  "married", "hisp", "re74", "re75", "u75", "u78")

#print table
df %>% kableExtra::kable() %>% kableExtra::kable_styling("striped")

```

I chose to match participants on unemployment in 1974, race, marital status, age, and level of education. While all covariates except ethnicity were significantly (marginally) different between individuals who did and did not receive treatment, some of the variables may be causally post to treatment assignment. For instance, after receiving the NSW program, an individual would likely have different employment status in 1975 and 1978, so I did not match/condition on these. Similarly, wages in 1974 and 1975 are likely to be causally post to treatment assignment (assuming 1974 wages are measured at the year's end and treatment is given near the beginning). Conversely, being unemployed in 1974 is a confounder, because it directly affects your likelihood of getting treatment AND later earnings (if you have gaps in employment, likely your earnings will be lower even 4 years down the road), and similar for race, marital status, age and education.
After matching on unemployment in 1974, race, marital status, age, and education, the average treatment effect of participation in the NSW program was a nonsignificant increase of \$1151.70 (se = \$1754.80). After matching, participants were not balanced on 1975 earnings, 1978 earnings, unemployment in 1975, and unemployment in 1978, but were balanced on other variables (however, the distributions of age and education differed by treatment status, even if means were balanced).

f.
```{r}
#f.
#use exact matching on education, race, ethnicity and married.

#make matches
match.mahF <- Match(Y = psid$re78,
                   Tr = psid$nsw,
                   X = psid[, c("hisp", "black", "educ", "married")],
                   M = 1, estimand="ATT",
                   Weight = 2,
                   exact = TRUE)
#print ATT
summary(match.mahF)

#check balance on all covariates (matched + unmatched)
mbF <- MatchBalance(nsw ~ age + educ + black + u74 +
             married + hisp + re74 + re75  +
             u75 + u78,
             data = psid, 
             match.out = match.mahF,
             nboots = 10)

#make a nicer looking balance table
#vector of covariates after matching
after.tr <- c(mbF$AfterMatching[[1]][3],
           mbF$AfterMatching[[2]][3],
           mbF$AfterMatching[[3]][3],
           mbF$AfterMatching[[4]][3],
           mbF$AfterMatching[[5]][3],
           mbF$AfterMatching[[6]][3],
           mbF$AfterMatching[[7]][3],
           mbF$AfterMatching[[8]][3],
           mbF$AfterMatching[[9]][3],
           mbF$AfterMatching[[10]][3])

#vector of values in control after matching
after.co <- c(mbF$AfterMatching[[1]][4],
           mbF$AfterMatching[[2]][4],
           mbF$AfterMatching[[3]][4],
           mbF$AfterMatching[[4]][4],
           mbF$AfterMatching[[5]][4],
           mbF$AfterMatching[[6]][4],
           mbF$AfterMatching[[7]][4],
           mbF$AfterMatching[[8]][4],
           mbF$AfterMatching[[9]][4],
           mbF$AfterMatching[[10]][4])

#sd in treated and control
var.tr <- c(mbF$AfterMatching[[1]][5],
           mbF$AfterMatching[[2]][5],
           mbF$AfterMatching[[3]][5],
           mbF$AfterMatching[[4]][5],
           mbF$AfterMatching[[5]][5],
           mbF$AfterMatching[[6]][5],
           mbF$AfterMatching[[7]][5],
           mbF$AfterMatching[[8]][5],
           mbF$AfterMatching[[9]][5],
           mbF$AfterMatching[[10]][5])

var.co <- c(mbF$AfterMatching[[1]][6],
           mbF$AfterMatching[[2]][6],
           mbF$AfterMatching[[3]][6],
           mbF$AfterMatching[[4]][6],
           mbF$AfterMatching[[5]][6],
           mbF$AfterMatching[[6]][6],
           mbF$AfterMatching[[7]][6],
           mbF$AfterMatching[[8]][6],
           mbF$AfterMatching[[9]][6],
           mbF$AfterMatching[[10]][6])

#make numeric and get sqrts
var.tr <- as.numeric(var.tr)
var.co <- as.numeric(var.co)

sd.tr <- sqrt(var.tr)
sd.co <- sqrt(var.co)


#vector of pvalues comparing groups
match.p <- c(mbF$AfterMatching[[1]][7],
           mbF$AfterMatching[[2]][7],
           mbF$AfterMatching[[3]][7],
           mbF$AfterMatching[[4]][7],
           mbF$AfterMatching[[5]][7],
           mbF$AfterMatching[[6]][7],
           mbF$AfterMatching[[7]][7],
           mbF$AfterMatching[[8]][7],
           mbF$AfterMatching[[9]][7],
           mbF$AfterMatching[[10]][7])

#also p values for KS tests
ks.pval2 <- c(mbF$AfterMatching[[1]][9][[1]][[1]][[1]],
             mbF$AfterMatching[[2]][9][[1]][[1]][[1]],
             "--", #black is categorical
             "--", #u74 is categorical
             "--", #married is categorical
             "--", #hisp is categorical
             mbF$AftereMatching[[7]][9][[1]][[1]][[1]], #re74
             mbF$AfterMatching[[8]][9][[1]][[1]][[1]], #re75
             "--", #u75 is categorical
             "--") #u78 is categorical

#replace 0 with "<0.001"
ks.pval2 <- str_replace(ks.pval2, pattern = "0", replacement ="< 0.001")


#print balance table with KS pvalues
df.F <- as.data.frame(cbind(sprintf(after.tr, fmt = "%.3f"),
                    sprintf(after.co, fmt = "%.3f"),
                    sprintf(sd.tr, fmt = "%.3f"),
                    sprintf(sd.co, fmt = "%.3f"),
                    sprintf(match.p, fmt = "%.3f"),
                    ks.pval2)) 
names(df.F) <- c("mean.tr", "mean.co",
               "sd.tr", "sd.co",
               "pvalue", "KSpvalue")

rownames(df.F) <- c("age", "educ", "black", "u74",
                  "married", "hisp", "re74", "re75", "u75", "u78")

df.F %>% kableExtra::kable() %>% kableExtra::kable_styling("striped")



```

The ATT using the exact matching procedure is a significant *decrease* of \$5821.60 (se = \$840.65). In this round of matching, there is balance on education, race, marital status, and ethnicity (what we matched on), but not on the other variables. The results are also different in that, among the matched variables, p values are =1 rather than simply > 0.05, reflecting the exact matching process. We have fewer observations that were able to be matched, and the estimate of the effect is in the opposite direction. The results differ both because of the different covariates that we matched on and because of the exact matching process: we threw away additional observations that didn't have exact matches, and limited the variation of the covariates rather than just reducing it.

g.

```{r}
#g. Re-estimate att with bias corrections

#match using M =1, bias corrected
m.biascorr1 <- Match(Y = psid$re78,
                     Tr = psid$nsw, 
                     X = psid[, c("hisp", "black", "educ", "married")],
                     M = 1, 
                     estimand="ATT",
                     Weight = 2,
                     BiasAdjust = TRUE)
summary(m.biascorr1)


#match using M =4, bias corrected
m.biascorr4 <- Match(Y = psid$re78,
                     Tr = psid$nsw, 
                     X = psid[, c("hisp", "black", "educ", "married")],
                     M = 4, 
                     estimand="ATT",
                     Weight = 2,
                     BiasAdjust = TRUE)
summary(m.biascorr4)


#match using M =10, bias corrected
m.biascorr10 <- Match(Y = psid$re78,
                     Tr = psid$nsw, 
                     X = psid[, c("hisp", "black", "educ", "married")],
                     M = 10, 
                     estimand="ATT",
                     Weight = 2,
                     BiasAdjust = TRUE)
summary(m.biascorr10)
##################################################
#without bias correction
#match using M =1, bias corrected
m.nocorr1 <- Match(Y = psid$re78,
                     Tr = psid$nsw, 
                     X = psid[, c("hisp", "black", "educ", "married")],
                     M = 1, 
                     estimand="ATT",
                     Weight = 2,
                     BiasAdjust = F)
summary(m.nocorr1)


#match using M =4, bias corrected
m.nocorr4 <- Match(Y = psid$re78,
                     Tr = psid$nsw, 
                     X = psid[, c("hisp", "black", "educ", "married")],
                     M = 4, 
                     estimand="ATT",
                     Weight = 2,
                     BiasAdjust = F)
summary(m.nocorr4)


#match using M =10, bias corrected
m.nocorr10 <- Match(Y = psid$re78,
                     Tr = psid$nsw, 
                     X = psid[, c("hisp", "black", "educ", "married")],
                     M = 10, 
                     estimand="ATT",
                     Weight = 2,
                     BiasAdjust = F)
summary(m.nocorr10)





```

The estimates in the bias-adjusted estimates are less extreme and standard errors larger than those that aren't bias adjusted. At the same time, as the number of matches increases, the number of observations and the magnitude of the effect both increase (i.e. the ATT gets more extreme, in this case more negative). The differences may be accounted by sample size and covariance between matching variables. I trust the 1:1 matching and the bias-adjusted matches more because matching 1:1 should, on average, make the matches more similar since we only match to the closest observation rather than including several that are less similar to the treated unit. The bias-adjustment should perform better than the uncorrected match because it accounts for confounders and specifically the conditonal expectation of those confounders/matching variables given other included matching variables.

h.

```{r}
#h. 
#match on the propensity score from part d
pscore.match <- Match(Y = psid$re78,
                      Tr = psid$nsw,
                      X = psid$pscore,
                      Weight = 2,
                      estimand = "ATT",
                      M = 1) 

summary(pscore.match)

```

When matching on propensity scores, the ATT is a nonsignificant increase of \$923.98 (se = \$1693.80)


i.
```{r}
#i.
#set seed
set.seed(01239)

#make function that can be bootstapped to get se of propensity score
calc_pscore <- function(){
  
  #take a random sample of rows from the data frame
  psid.sub <- psid[sample(1:nrow(psid), size = 1000, replace = F),]

  #estimate the propensity scores
  ps.mod <- glm(nsw ~ hisp + black + educ + married,
                data = psid.sub)
  
  #get the scores
  psid.sub$ps.scores <- predict(ps.mod, type = "response")
  
  #define the PS weights for control group to get the ATT
  psid.sub$ps.weight <- ifelse(psid.sub$nsw == 0, 
                               (psid.sub$ps.scores/(1 - psid.sub$ps.scores)), 0)
  
  #get the ATT
  ATT <- (1/sum(psid.sub$nsw == 1))*
    sum((psid.sub$nsw * psid.sub$re78) - (psid.sub$re78*(1 - psid.sub$nsw)* psid.sub$ps.weight), na.rm = T)
  
  #return the ATT
  return(ATT)
}

#bootstap the estimate
boot.ps <- replicate(n = 10000, expr = calc_pscore(), simplify = "vector")

#take a look at it
hist(boot.ps)

#get estimate and standard error of the estimate
mean(boot.ps)
sqrt(var(boot.ps))

```

Using the bootstrapped sample, the estimated effect of treatment on the treated is a decrease of \$6970 (se = \$996.58).


j.
```{r}
#j.
#use genmatch


#The covariates we want to obtain balance on
BalanceMat <- cbind(psid$age, psid$educ, psid$black, psid$hisp, 
                    psid$married, psid$u74, psid$u75, psid$re75,
                    psid$re74)

#get weight matrix
genout <- GenMatch(Tr=psid$nsw,
                   X=psid[, c("hisp", "black", "educ", "married",
                              "u74", "u74", "re75", "re74", "age")],
                   BalanceMatrix = BalanceMat,
                   estimand="ATT",
                   M=1,
                   pop.size=1000,
                   max.generations=10,
                   wait.generations=1)

#match
match.gen <- Match(Y=psid$re78,
                   Tr=psid$nsw,
                   X=psid[, c("hisp", "black", "educ", "married",
                              "u74", "u74", "re75", "re74", "age")],
                   estimand="ATT",
                   Weight.matrix=genout)
summary(match.gen)

#check balance
MatchBalance(nsw ~ age + educ + black + u74 +
             married + hisp + re74 + re75  +
             u75 + u78,
             data = psid, 
             match.out = match.gen,
             nboots = 10)


```

Using genetic matching, the ATT is estimated to be a nonsignificant increase of \$2248.1 (se = \$1784.30). All covariates are balanced except age, education, and unemployment in 1978. While it takes longer, I prefer the genetic matching because of the iterative and automated matching process, and the ability to acheive good balance (at least using this data).


k.

The ATT is identified if there are no open backdoor paths confounding the association between program participation and 1978 earnings (including those by unobserved confounders), if we have not conditioned on any colliders, if there is no measurement error, and if we haven't conditioned on post-treatment covariates. Matching makes assumptions about confounding/blocking backdoor paths somewhat more plausible, but, as in regression, the success of matching in controlling for confounding depends on appropriate selection of matched covariates. We also cannot match on unobserved covariates/confounders, which may still differ between treatment and control groups after matching; thus, omitted variable bias is still possible. However, matching can be helpful in ensuring balance on what you have measured and positivity, whereas regression can violate positivity if there are no/few observations with given combinations of covariates.


